The Digital Edition of the Archaic Latin Inscriptions (7th{}-5th century B.C.)

### 2016-01-18 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
Paper 22 (Sarullo et al) edited up to end of Section 1. NOT FINISHED!

\maketitle
\begin{abstract}
This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
EpiDoc Guidelines. The edition is the result of an autoptical examination of the epigraphic documents and of the
text-bearing objects, together with the analysis of previous studies. In the particular case of the Forum inscription,
this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § 1 and 2;
Giulia Sarullo (GS) is responsible for § 4 and Marta Muscariello (MM) for § 3.}
\end{abstract}
\keywords{Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.}

\section{Generalia [GR]}

\noindent The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
Latium\footnote{With the two exceptions of the Vendia's Urn, found in Cerveteri, and the Garigliano bowl.}
dating back to the period between the 7th and the 5th century B.C. Between the end of the 19th century and the first
decades of the 20th century this corpus consisted of only four inscriptions of a certain length, that is the Duenos
vase (1880), the Fibula Praenestina (1887), the Forum inscription (1899) and the Tibur pedestal inscription (1926),
besides other shorter but still very interesting texts that offered important cultural information (such as the
inscription from the Regia, REX, CIL I2 479). In the second half of the 20th century, the corpus grew significantly and
reached the total of about eighty documents that have not been gathered in a comprehensive edition yet. The updated
corpus includes new and crucial discoveries, some of which have already been published in a traditional way (in print),
such as the inscriptions from Satricum and the fragments found at the Palatinum, whereas others are still unpublished,
as in the case of the recent findings from the Regia. The quantitative and qualitative enrichment provided by recent
new readings of the texts and the new data emerged justify and require a new publication.

Most of the inscriptions in the collection are frustuli or single letters: these texts, though not really relevant from
the linguistic point of view, are important testimony of the use of writing in Latium since the 7th century B.C.

The website is then an absolute novelty in the field of digital epigraphy, since at the moment no online epigraphic
collection specifically dedicated to these documents exists. It is well known that, for every kind of publication, a
digital edition provides several advantages, for example: the possibility to update both the textual corpus and the
bibliography continuously; the hypertextual structure, that allows the user to utilize the edition in different ways;
the opening to a public that is heterogeneous and wider than the one a work addressed to specialists can reach. In
which other ways can the web respond to the specific requirements of a peculiar corpus such as ours with distinguishing
features, different from the later Latin epigraphy optimally reproduced in the Epigraphic Database Rome (EDR)?

First of all, by using the EpiDoc encoding standard, which is compatible with other encoding systems, it will be
possible to transfer our data to EDR and to the EAGLE-Europeana Network (see § 4).

@@

Besides what has been illustrated so far, the presence of complex indexes, through which it will be possible to locate
the inscription from different starting points, will allow the user to search the texts according to various
parameters: dating, place of finding, object type, and textual type (see § 3).

The choice of the method to adopt as regards the interpretative transcription is problematic: it is clear that it is
absolutely impossible to be completely neutral or objective. Unlike the diplomatic transcription, for which a critical
apparatus can be constructed and be extremely useful, the same is not possible for the interpretative transcription: an
exhaustive critical apparatus would imply a superabundance of information that could negatively influence the
scientific nature and the usability of the edition, especially in the case of inscriptions that have been variously
interpreted since their discovery. For longer inscriptions (e.g. Forum Inscription, Duenos vase, Tivoli inscription,
Garigliano bowl), the numerous readings that have been proposed so far by scholars have been compared and verified in
the light of linguistic criteria and of the new data in order to obtain an edition of the text that, although it cannot
be considered the definitive interpretation and does not solve all the pending issues, poses itself as a new starting
point for future research. A cross-reference to all the other interpretations will offer a complete source of
information and a tool that intends to be useful and exhaustive.

The archaic Latin inscriptions play a fundamental role in the study of the first stages of the language, since they
present particular features that allow us to investigate the various steps that led to “standard Latin”. The language
attested in our inscriptions can be considered a Restsprache insofar as it is not `readable' through later Latin
but it can be `interpreted'. The linguistic commentary will be carried out, in votis, in the second phase of this
project. Here the research focuses on the epigraphic features, that show a plurality of forms and of alphabets in such
a limited corpus. The chronological and geographical distribution of the signs and of the variants in use in the
inscriptions have been analyzed in order to offer valid elements to the study of the evolution of the alphabetic model
between the 7th and the 5th century B.C.

\section{Epigraphic News about the Forum Inscription [GR]}

@@

\section{The Epigraphic Chart [MM]}

@@

\section{Technical notes [GS]}

>>>>

\maketitle
\begin{abstract}
This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
EpiDoc Guidelines. The edition is the result of an autoptical examination of the epigraphic documents and of the
text-bearing objects, together with the analysis of previous studies. In the particular case of the Forum inscription,
this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for Sec. \ref{sec:generalia} and \ref{sec:news};
Giulia Sarullo (GS) is responsible for Sec. \ref{sec:3chart} and Marta Muscariello (MM) for Sec. \ref{sec:4tech}.}
\end{abstract}
\keywords{Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.}

\section{\emph{Generalia} [GR]}\label{sec:generalia}

\noindent The project \emph{Iscrizioni Latine Arcaiche (ILA)} consists in the digital edition of the inscriptions found in old
\emph{Latium}\footnote{With the two exceptions of the \emph{Vendia's} Urn, found in Cerveteri, and the Garigliano bowl.}
dating back to the period between the 7th and the 5th century B.C. Between the end of the 19th century and the first
decades of the 20th century this \emph{corpus} consisted of only four inscriptions of a certain length, that is the \emph{Duenos}
vase (1880), the \emph{Fibula Praenestina} (1887), the \emph{Forum} inscription (1899) and the \emph{Tibur} pedestal inscription (1926),
besides other shorter but still very interesting texts that offered important cultural information (such as the
inscription from the \emph{Regia, REX, CIL} I$^2$ 479). In the second half of the 20th century, the \emph{corpus} grew significantly and
reached the total of about eighty documents that have not been gathered in a comprehensive edition yet. The updated
\emph{corpus} includes new and crucial discoveries, some of which have already been published in a traditional way (in print),
such as the inscriptions from \emph{Satricum} and the fragments found at the \emph{Palatinum}, whereas others are still unpublished,
as in the case of the recent findings from the \emph{Regia}. The quantitative and qualitative enrichment provided by recent
new readings of the texts and the new data emerged justify and require a new publication.

Most of the inscriptions in the collection are \emph{frustuli} or single letters: these texts, though not really relevant from
the linguistic point of view, are important testimony of the use of writing in \emph{Latium} since the 7th century B.C.

The website is then an absolute novelty in the field of digital epigraphy, since at the moment no online epigraphic
collection specifically dedicated to these documents exists. It is well known that, for every kind of publication, a
digital edition provides several advantages, for example: the possibility to update both the textual \emph{corpus} and the
bibliography continuously; the hypertextual structure, that allows the user to utilize the edition in different ways;
the opening to a public that is heterogeneous and wider than the one a work addressed to specialists can reach. In
which other ways can the web respond to the specific requirements of a peculiar \emph{corpus} such as ours with distinguishing
features, different from the later Latin epigraphy optimally reproduced in the \emph{Epigraphic Database Rome} (EDR)?

First of all, by using the EpiDoc encoding standard, which is compatible with other encoding systems, it will be
possible to transfer our data to EDR and to the EAGLE-Europeana Network (see Sec. \ref{sec:4tech}).

@@

Besides what has been illustrated so far, the presence of complex indexes, through which it will be possible to locate
the inscription from different starting points, will allow the user to search the texts according to various
parameters: dating, place of finding, object type, and textual type (see Sec. \ref{sec:3chart}).

The choice of the method to adopt as regards the interpretative transcription is problematic: it is clear that it is
absolutely impossible to be completely neutral or objective. Unlike the diplomatic transcription, for which a critical
apparatus can be constructed and be extremely useful, the same is not possible for the interpretative transcription: an
exhaustive critical apparatus would imply a superabundance of information that could negatively influence the
scientific nature and the usability of the edition, especially in the case of inscriptions that have been variously
interpreted since their discovery. For longer inscriptions (\emph{e.g.} \emph{Forum} Inscription, \emph{Duenos} vase, Tivoli inscription,
Garigliano bowl), the numerous readings that have been proposed so far by scholars have been compared and verified in
the light of linguistic criteria and of the new data in order to obtain an edition of the text that, although it cannot
be considered the definitive interpretation and does not solve all the pending issues, poses itself as a new starting
point for future research. A cross-reference to all the other interpretations will offer a complete source of
information and a tool that intends to be useful and exhaustive.

The archaic Latin inscriptions play a fundamental role in the study of the first stages of the language, since they
present particular features that allow us to investigate the various steps that led to ``standard Latin''. The language
attested in our inscriptions can be considered a \emph{Restsprache} insofar as it is not `readable' through later Latin
but it can be `interpreted'. The linguistic commentary will be carried out, \emph{in votis}, in the second phase of this
project. Here the research focuses on the epigraphic features, that show a plurality of forms and of alphabets in such
a limited \emph{corpus}. The chronological and geographical distribution of the signs and of the variants in use in the
inscriptions have been analyzed in order to offer valid elements to the study of the evolution of the alphabetic model
between the 7th and the 5th century B.C.

\section{Epigraphic News about the Forum Inscription [GR]}\label{sec:news}

@@

\section{The Epigraphic Chart [MM]}\label{sec:3chart}

@@

\section{Technical notes [GS]}\label{sec:4tech}
